---
title: 'Assignment 5: Conditional Variance and Local Poisson'
author: "Víctor Villegas, Roger Llorenç, Luis Sierra"
date: "2024-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Conditional Variance

Loading the necessary library and dataset. Storing the logarithm transform of the variables.

```{r}
library(sm)
data(aircraft)
help(aircraft)
attach(aircraft)
lgPower <- log(Power)
lgSpan <- log(Span)
lgLength <- log(Length)
lgWeight <- log(Weight)
lgSpeed <- log(Speed)
lgRange <- log(Range)
```

We consider a heteroscedastic regression model $Y = m(X) + \sigma(X)\varepsilon$ for $\varepsilon$ the standard, zero-mean Gaussian noise.

```{r}
plot(Yr, lgWeight)
```

## 1.1. Nonparametric regression model on the original data

```{r}
# Function loc.pol.reg option
source("locpolreg.R")
aircraft.lp_reg <- locpolreg(Yr, lgWeight)
```

```{r}
# Function sm.regression option
library(KernSmooth)
h <- dpill(Yr, lgWeight)
aircraft.sm_reg <- sm.regression(Yr, lgWeight, h)
```

## 1.2. Transformed estimated residuals

```{r}

```

# 2. Local Poisson Regression

## 2.1. Bandwidth choice

## 2.2. Local Poisson regression for Country Development

```{r}
data <- read.csv2('HDI.2017.subset.csv')
```

