---
title: 'Assignment 5: Conditional Variance and Local Poisson'
author: "Víctor Villegas, Roger Llorenç, Luis Sierra"
date: "2024-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Conditional Variance

Loading the necessary library and dataset. Storing the logarithm transform of the variables.

```{r}

# Clears plots
while (dev.cur() != 1) {
  dev.off()
}
# Clears global environment
rm(list=ls())


library(sm)
data(aircraft)
help(aircraft)
attach(aircraft)
lgPower <- log(Power)
lgSpan <- log(Span)
lgLength <- log(Length)
lgWeight <- log(Weight)
lgSpeed <- log(Speed)
lgRange <- log(Range)
```

We consider a heteroscedastic regression model $Y = m(X) + \sigma(X)\varepsilon$ for $\varepsilon$ the standard, zero-mean Gaussian noise.

```{r}
plot(Yr, lgWeight)
```

## 1.1. Nonparametric regression model on the original data

```{r}
# Function loc.pol.reg option
source("locpolreg.R")

# Leave-one-out CV to select bandwidth
h.cv.gcv <- function(x,y,h.v = exp(seq(log(diff(range(x))/20),
                                       log(diff(range(x))/4),l=10)), 
                     p=1,type.kernel="normal"){
  n <- length(x)
  cv <- h.v*0
  gcv <- h.v*0
  for (i in (1:length(h.v))){
    h <- h.v[i]
    aux <- locpolreg(x=x,y=y,h=h,p=p,tg=x,
                     type.kernel=type.kernel, doing.plot=FALSE)
    S <- aux$S
    h.y <- aux$mtgr
    hii <- diag(S)
    av.hii <- mean(hii)
    cv[i] <- sum(((y-h.y)/(1-hii))^2)/n
    gcv[i] <- sum(((y-h.y)/(1-av.hii))^2)/n
  }
  return(list(h.v=h.v,cv=cv,gcv=gcv))
}
```

```{r}

h.v <-  exp(seq(from=log(1), to = log(20), length=30))
out.h.cv <- h.cv.gcv(x=aircraft$Yr, y=lgWeight, h.v=h.v)
h.loo.cv <- h.v[which.min(out.h.cv$cv)]

```

```{r}

plot(h.v,out.h.cv$cv, xlab ="bandwidth", ylab = "mean ssr")
lines(h.v,out.h.cv$cv)
abline(v = h.loo.cv, col = "red") 

aircraft.lp_reg <-locpolreg(x=aircraft$Yr,y=lgWeight,h=h.loo.cv)

```

```{r}
# Function sm.regression option
library(KernSmooth)

h2 <- dpill(x=aircraft$Yr,y=lgWeight,gridsize=length(aircraft$Yr),range.x=range(aircraft$Yr))
set.seed(123)
aircraft.sm_reg <- sm.regression(x = aircraft$Yr, y = lgWeight, h = h2, eval.points = seq(min(aircraft$Yr), max(aircraft$Yr), length.out = length(aircraft$Yr)))

```

## 1.2. Transformed estimated residuals

```{r}
# Function loc.pol.reg option
residualss1 <- lgWeight - aircraft.lp_reg$mtgr
z_i1 = log(residualss1^2)
h.v_z <-  exp(seq(from=log(1), to = log(30), length=30))

out.h.cv_z <- h.cv.gcv(x=aircraft$Yr, y=z_i1, h.v=h.v_z)
h.loo.cv_z <- h.v[which.min(out.h.cv_z$cv)]

```


```{r}
plot(h.v_z,out.h.cv_z$cv, xlab ="bandwidht", ylab = "mean ssr")
lines(h.v_z,out.h.cv_z$cv)
abline(v = h.loo.cv_z, col = "red") 
```

```{r}
q_hat1 = locpolreg(x=aircraft$Yr,y = z_i1,h = h.loo.cv_z)
```


```{r}

sigma_square_hat1 = exp(q_hat1$mtgr)


plot(aircraft$Yr, residualss1^2, col = rgb(0, 0, 0, alpha = 0.4), xlab = "x", ylab = "e_i^2")
lines(aircraft$Yr, sigma_square_hat1, col = "red", lwd = 3)  # Adjust line width here

# Your plot code
plot_aux <- locpolreg(x=aircraft$Yr,y=lgWeight,h=h.loo.cv)
lines(aircraft$Yr, aircraft.lp_reg$mtgr, type = "l", col = "blue")
lines(aircraft$Yr, aircraft.lp_reg$mtgr + 1.96 * sqrt(sigma_square_hat1), col = "red")
lines(aircraft$Yr, aircraft.lp_reg$mtgr - 1.96 * sqrt(sigma_square_hat1), col = "red")
grid(col = "gray", lty = "dotted")  # Add grid

# Add legend with adjusted parameters
legend("topleft", legend = c("Data", "m_hat", "Confidence Interval"), 
       col = c(rgb(0, 0, 0, alpha = 0.5), "blue", "red"), lty = c(NA, 1, 1), 
       pch = c(1, NA, NA), cex = 0.6)

```

```{r}
# Function sm.regression option
residualss2 <- lgWeight - aircraft.sm_reg$estimate
z_i2 = log(residualss2^2)

h.dpill_zi2 <- dpill(x=aircraft$Yr,y=z_i2,gridsize=length(aircraft$Yr),range.x=range(aircraft$Yr))

q_hat2 = sm.regression(x = aircraft$Yr, y = z_i2, h = h.dpill_zi2, eval.points = seq(min(aircraft$Yr), max(aircraft$Yr), length.out = length(aircraft$Yr)))

sigma_square_hat2 = exp(q_hat2$estimate)


plot(aircraft$Yr, residualss2^2, col = rgb(0, 0, 0, alpha = 0.4), xlab = "x", ylab = "e_i^2")
lines(aircraft$Yr, sigma_square_hat2, col = "red", lwd = 3)  # Adjust line width here

# Your plot code
set.seed(123)
plot_aux_2 <- sm.regression(x = aircraft$Yr, y = lgWeight, h = h2, eval.points = seq(min(aircraft$Yr), max(aircraft$Yr), length.out = length(aircraft$Yr)))
# plot(aircraft$Yr, lgWeight, col = rgb(0, 0, 0, alpha = 0.5), xlab = "x", ylab = "m_hat")
lines(aircraft$Yr, aircraft.sm_reg$estimate, type = "l", col = "blue")
lines(aircraft$Yr, aircraft.sm_reg$estimate + 1.96 * sqrt(sigma_square_hat2), col = "red")
lines(aircraft$Yr, aircraft.sm_reg$estimate - 1.96 * sqrt(sigma_square_hat2), col = "red")
grid(col = "gray", lty = "dotted")  # Add grid

# Add legend with adjusted parameters
legend("topleft", legend = c("Data", "m_hat", "Confidence Interval"), 
       col = c(rgb(0, 0, 0, alpha = 0.5), "blue", "red"), lty = c(NA, 1, 1), 
       pch = c(1, NA, NA), cex = 0.6)


```


# 2. Local Poisson Regression

## 2.1. Bandwidth choice

## 2.2. Local Poisson regression for Country Development

```{r}
data <- read.csv2('HDI.2017.subset.csv')
```

